---
title: "A codebook for summarizing presence or absence of stone tool-making techniques reported in the literature"
author: -Jonathan Paige^[Institute of Human Origins, School of Human Evolution and
  Social Change, Arizona State University, jnpaige@asu.edu]
bibliography: ["references.bib"] #This is the reference file in the rmarkdown directory. 
#https://github.com/crsh/citr more info on bib connection ot zotero here. 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    keep_tex: yes  #helps with linkage between latex and rmarkdown
    number_sections: yes
    toc: true
    toc_depth: 3
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{float}  #this helps with figure placement
- \floatplacement{figure}{hp}
- \usepackage{booktabs}  
- \usepackage[font=small,labelfont=bf]{caption}
- \usepackage{setspace}\doublespacing
#- \usepackage{dcolumn} #for rotation of stargazer tables. https://stackoverflow.com/questions/21840878/rotate-a-table-from-r-markdown-in-pdf
#always_allow_html: yes
#indent: true
abstract: |
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---


```{r, include=FALSE}
options(tinytex.verbose = TRUE)
#https://yihui.org/tinytex/r/#debugging
#https://www.cambridge.org/core/blog/2018/07/30/how-researchers-can-solve-the-bottle-opener-problem-with-compute-capsules/
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```
# Version details:

This report was generated on `r Sys.time()`
The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

# Introduction


  If an archaeologist wants to study more than a handfull of archaeological assemblages, or compare the material culture of several different cultures, they have two choices. Study all the relevant material themselves, or rely on published descriptions and data. Below I argue that we should be doing more of the latter, and that collecting data from existing studies helps us better realize the many benefits of the comparative method. However, some very real problems have stood in the way. Existing descriptions and data describing material culture are as much themselves texts as the original materials themselves (Go fucking post-processualist hogwild here). Any data table, prose summary, or set of illustrations describing material culture condenses those products into a kind of texts that themselves must be interpreted [@hodderInterpretationDocumentsMaterial1994].
  
Outline
  -
  
Comparative research. 


  Stone tool technologies are some of our only evidence for behavioral change across hominin evolution, and they hold great promise in improving our understanding of how humans evolved. This paper focuses on helping researchers to perform reproducible research in comparative stone tool analysis, which necessarily involves analyisis of data collected from many individuals. As others have written about at length, this is difficult to accomplish (Will et al. 2019, Riede et al. 2020). Below, I describe some of the problems facing comparative stone tool analysis, and some proposed solutions to give a broader understanding of these issues. However, the bulk of this paper focuses on outlining how the use of codebooks can help to ensure that comparative studies are reproducable.
  
  through use of carefully designed code-books. Code-books address one problem, namely, ensuring that any two researchers come to the same conclusions about how to measure something, or when to interpret some feature as present or absent. 
  
  
  But, to do sciennce, we need the desciption of the methods to convey in high enough detail for people to know specifically 
  
# Problems. Standardization in ontologies often is region, and time specific. 

In order to make conclusions about prehistory that we can be confident in, artifact classifications, and measurements must be reliable, and replicable. Any archaeological assemblage has a great deal of complex information that an archaeologist could attend to. Researchers might vary in their personal recording tendencies, and they might differ from one another as a result of belonging to different research traditions. As a consequence, inter-analyst variation, and error all hinder systematic studies of lithic technology (Gnaden and Holdoway, Fish 1978, Prentiss 1998, Whittaker et al. 1998, Beck and Jones 1989, Hodson 1966). For our research to be replicable, future researchers should be able to know what we measured, or how we classified material, or what standards we used for inferring certain kinds of behaviors, and why, without having to ask.

All sets of debitage and tools represent accumulations of many behaviors that likely occurred across generations. So, the number of procedural units and modes we identify could be driven in part by the recycle, re-use, and transportation of artifacts in prehistory, rather than reflecting a single coherent technological strategy (Coco et al., 2020; Rezek et al., 2020). When sites are discovered and excavated, there is further loss in artifacts whose size does not meet the collection threshold (screen size), which can obscure the presence of some kinds of core management techniques, and methods of producing tools (Towner & Warburton, 1990). During excavation and analysis, the chaotic accumulations of many behaviors are grouped into discrete assemblages, typically by major stratigraphic unit. These assemblages are then studied, and characterized (Rezek et al., 2020). During analysis, a researcher could misreport the presence or the absence of a manufacturing step, or inadvertently lump different manufacturing steps that belong to two or more independent reduction sequences because of site disturbance and time-averaging (Perreault, 2018). Such inter-observer variation and errors in reconstructing reduction sequences are to be expected (Beck & Jones, 1989; Fish, 1978; Gnaden & Holdaway, 2000; Hodson et al., 1966; Prentiss, 1998; Whittaker et al., 1998). Mechanical and chemical processes can further complicate these reconstructions. Overly patinated, rolled and worn pieces may have evidence of some behaviors erased, such as abrasion of platforms. These processes may also complicate attempts to identify how tools were retouched (Adler et al., 2014).
 	A lack of standardization, and common practice is especially an obstacle to analyses of lithic variation across sites, regions, and continents, where researchers who studied the assemblages to be included came from different research traditions. Benefits of comparative studies in lithic technology, and shared units of analysis. '


Another problem is that even in categories with explicit descriptions, and good reference illustrations, 
  -New forms accomodated under Yuma type. Catchall for non-clovis lanceolate points in early days of Paleoindian archaeology [@bousmanPaleoindianArcheologyTexas2004]
  -Case of new forms accomodated in Plainview type in 1954. 
  -Decades untill this was walked back [@bousmanPaleoindianArcheologyTexas2004]


	

 1. lack of standardization in how stone tool assemblages are studied.There are diverse approaches to stone tool research, driven by different kinds of questions, and with different theoretical perspectives, all of which have evolved over time. This means that it is difficult to compare many assemblages, especially between the products of distinct research traditions.   


2. lack of standardization in how the same kind of trait is measured and varying understanding of the definition of that trait, and

 The second problem is that individuals often assume that there is standardization in some measurement (like platform angle), or trait (what constitutes a prepared core), when there is actually between-analyst error in measuring, and variation in researchers understanding of the definition of those traits. This is probably less of a problem within research teams, or within research traditions. 
  

This is where there has been the most work in exploring inter-observer error. 

Coding errors solved by *enterer trois* E4, E5 data entry programs.




Any archaeological assemblage has a great deal of complex information that an archaeologist could attend to. Researchers might vary in their personal recording tendencies, and they might differ from one another as a result of belonging to different research traditions. As a consequence, inter-analyst variation, and error all hinder systematic studies of lithic technology (Gnaden and Holdoway, Fish 1978, Prentiss 1998, Whittaker et al. 1998, Beck and Jones 1989, Hodson 1966).  Research teams A and B might Some researchers might have a slightly different understanding of how to measure platform angle, and do not explicitly describe that understanding in papers. We know that stone tool researchers face problems in consistently performing those measurements.     


# Problems. High cost of re-analyzing assemblages. 

3, the huge cost of re-analyzing assemblages once a standardized method of analysis is developed. 

  Thirdly, systematically analyzing each artifact in even one assemblage of moderate size is a long, time-consuming task. For example, the analysis of 15,000 stone tools at the Pinnacle Point 5-6 site involved the efforts of 5 expert lithic analysts working every day together for weeks. For my own dissertation research, I budgeted ~14 10 hour work days for each of the assemblages I sampled and analyzed. A fuller analysis of each site's assemblage in some cases would take months. The cost of analyzing an assemblage, means that for research questions about broad changes in technology at the continental, and hundred-thousand year to million year scale, cannot be achieved by re-analyzing collections previously reported. Furthermore, travelling overseas to analyze multiple assemblages has a huge carbon footprint. Requires large budgets to perform research, which may not be available to many early career researchers and students. Also, its not covid friendly. 
  

Financial cost

Time Cost

Carbon emissions

Feasibility of travel in post-covid world


# Summary



 	A lack of standardization, and common practice is especially an obstacle to analyses of lithic variation across sites, regions, and continents, where researchers who studied the assemblages to be included came from different research traditions. Benefits of comparative studies in lithic technology, and shared units of analysis. 
	
	Rezek et al. 
Will et al. 
	Riede, Hoggard, Shennan.
	
Overall, it is likely that some kinds of measurements and ways of making sense of the complex information that can be gleaned from stone tool assemblages are far more reliable than others (e.g. measurements of the maximum length of a flake). Even simple, reliable measurements serve as the basis for very important broad comparative studies (Rezek et al. 2018). However, much of the information we are interested in relates to more complex information about how cores are managed, changes in technologies over time beyond a narrow set of artifact measurements, or across sets of artifact classes as opposed to within a single artifact class. When we begin to try to operationalize change in this more complex domain of technological variation, we have run into issues. 


# Solutions


  Below I discuss these problems a bit more in depth, 
  
  
  Solutions proposed by archaeologists. 
	Record the learning pedigrees of lithicists in order to better understand biases (). 
	Explicit definitions of types, attributes, etc. 
		Whittaker et al. 1998. 
Calibration of workers to a more experienced analyst. Double checking of data by experienced analyst (Gnadena and Holdaway 2000). 
Working in large teams of well-trained researchers so that things can be double checked to improve reliability (Wilkins et al. 2017)
	Computational methods of comparison (Grosman et al. 2008). 

	Reliance on type collections. 	

  
  
  However, comparative studies of technological change face several challenges. 
  One solution to this problem has been to collect data on stone tool assemblages that relate to relatively coarse units of analysis, and can be recovered from the literature. For example, outs
  
  
  
  
  
  
  Just as stone tool research approaches have evolved, so have the methods of how stone tool assemblages are analyzed and reported. Even within the African Middle Stone Age, a somewhat focused field of study, there is little agreement about what should be measured, and reported (Tryon and Will).  
 








  More to the point, the standards by which some traits are argued to be present, or the reasoning behind argued classifications, are often implicit.
  
  Because classifications are implicit, standardization is not possible, and attempts at universal methods of measuring variability are unlikely to be accepted. 
  A lack of standardization, and common practice is especially an obstacle to analyses of lithic variation across sites, regions, and continents. Benefits of comparative studies in lithic technology, and shared units of analysis. 
	
	Rezek et al. 
Will et al. 
	Riede, Hoggard, Shennan.
	

Overall, it is likely that some kinds of measurements and ways of making sense of the complex information that can be gleaned from stone tool assemblages are far more reliable than others (e.g. measurements of the maximum length of a flake). Even simple, reliable measurements serve as the basis for very important broad comparative studies (Rezek et al. 2018). However, much of the information we are interested in relates to more complex information about how cores are managed, changes in technologies over time beyond a narrow set of artifact measurements, or across sets of artifact classes as opposed to within a single artifact class. When we begin to try to operationalize change in this more complex domain of technological variation, we have run into issues. 

Stone tool research itself is made up of individuals, who each may have different experiences and tendencies when it comes to how they classify and report variation at archaeological sites. These differences cause error in coding the presence or absence of different kinds of artifact attributes (Gnaden and Holdaway 2000)., and on differences in how artifacts are fit into a given typology (citations in Gnaden and Holdaway 2000:741 P1). 

Artifact descriptions as well may include terms that, over time, could reasonably be interpreted to cover a greater diversity of artifact features than any one individual who originally defined the term might anticipate (Gilboa et al. 2004:682). In this sense, how we publish descriptions of artifact types, technological industries, is somewhat similar to how definitions of words are recorded in English dictionaries. Dictionaries do not place restrictions on what meanings a word can have. So, words can accumulate new meanings as semantics evolve (Traugott and Dasher 2001). As a given word acquires new meanings in its usage, those can be added to the dictionary, either as an additional meaning, or as a homonym. This affords significant flexibility to how language develops. A word only must only enjoy common usage to enter the vernacular, and there are few mechanisms imposed by dictionaries to restrict what a word can mean. Those mechanisms largely operate in the social realm, as words are imitated, and taught, co-opted, experimented with, or misunderstood in various contexts.
The flexibility of language is a problem for certain kinds of scientific research. For example, epidemiologists who study the efficacy of different kinds of health interventions on groups often rely on analyzing interview transcripts where an individual self report.. Perhaps a health intervention could make someone healthier, or less healthy,  make no difference either way, or perhaps there isn’t enough information in the interview to make any sort of conclusion. Often the context of a statement matter a great deal “Never been better” stated sarcastically in the middle of a hacking cough means something very different from being stated after a contented sigh. Given the flexility in which people communicate, researchers need explicit and unambiguous standards for when to make the determination between those four options based on the interview transcript. How much are the coders allowed to read-between-the-the-lines of such an interview?  Should they take the context into account, or do we always take “never been better” as evidence of the intervention working? If it is ever ambiguous like this do opt not to make a judgement call and remove that datapoint?  Researchers need to make sure that the means of deciding between how to code something like this aren’t in the individual coder, but are based on standardized criteria. Often, coders do come to very different conclusions about the contents of an interview (Hruscka et al., Macqueen et al.) and this is a big problem in reproducibility of research, and our ability to identify patterns. 

In situations where there is complex information, and someone needs to make a decision, they are liable to fall on heuristics, and prior knowledge in making that judgement (Tversky and. Hruscka et al. macqueen et al.). This is one way in which systematic bias can be introduced to a study. 

Solutions proposed by archaeologists. 
	Record the learning pedigrees of lithicists in order to better understand biases (). 
	Explicit definitions of types, attributes, etc. 
		Whittaker et al. 1998. 
Calibration of workers to a more experienced analyst. Double checking of data by experienced analyst (Gnadena and Holdaway 2000). 
Working in large teams of well-trained researchers so that things can be double checked to improve reliability (Wilkins et al. 2017)
	Computational methods of comparison (Grosman et al. 2008). 
	Reliance on type collections. 	


## Solutions outside archaeology

Other fields in cognitive science, epidemiology, and medicine, have faced analogous problems, and have engaged in decades of development of best practices for breaking down massively complex data into coded descriptions that are reliable, and serve to answer questions about some phenomenon. People often rely on rules of thumb, and intuitive judgements when processing complex information of any kind (Hruschka et al. 2004, Kahneman et al. 1982). The more difficult that information is to process, the more inter-observer reliability is likely to suffer (Hruschka et al. 2004, Hagelin 2009). While working in teams, and calibrating workers to a single analyst are effective, they do not solve the problem in the long term. 

To improve inter-observer reliability, we have developed a codebook following the design of those used at the Center for Disease Control for analyzing interview transcripts (MacQueen et al. 1998). Codebook entries include short and long descriptions of the procedural unit, typical and atypical examples as well as examples that are close, but are not to be coded as the procedural unit in question. Each example includes a mix of written terms and phrases as well as illustrations that might be found in a text (Appendix). While the codebook ensures the standards by which we count a procedural unit as present or absent is replicable, it cannot control for differences in how stone tool assemblages are reported. For example, typological summaries of lithic technologies report very different information from in depth technological summaries of lithic assemblages. To address this, we will sample only from unusually detailed technological summaries of lithic assemblages. Studies that report Levallois as present, for example, but do not offer detailed descriptions of how convexities were managed, and how platforms were prepared, were not sampled.  For this analysis, we also targeted the most complex chain reduction sequences present at each site (i.e. this is not a census of all techniques present at a site, just within a given tool making sequence). This is because we are not necessarily interested in the central tendency in complexity over time, but in when new thresholds were breached.  In total, we gathered and analyze procedural unit data on 52 tool making sequences. 

I made this codebook while working on ways of quantifying variation in tool-making sequences across the archaeological record. In this case the unit of analysis is the technique, or procedural unit (Perreault et al., 2013). I coded procedural units as present or absent across archaeological assemblages, based on published reports describing those assemblages. Most systematic, comparative work on lithic technology has focused on units of analysis that are more consistent, and replicable across assemblages. For example flake and biface measurements, and shape variation can be very consistent between studies. Other aspects of lithic technology, like chaînes opératoire or reduction sequences that we are interested in, are not as easy to analyze statistically. Similarity and dissimilarity in them is difficult to quantify systematically because they represent qualitative data conveyed through prose descriptions, tables of artifact counts, and illustrations of individual artifacts and schematics of those sequences. How sequences are reported might vary by the research tradition of the analyst, and how I interpret those reports is also shaped by the tradition I am a part of. This is a big hurdle to making systematic comparisons of technology across many assemblages (Reynolds, 2018). 

Reliably extracting the presence or absence of procedural units from the complex descriptions of stone tool technologies is a similar problem to that faced by scientists who code interview transcripts, and other texts. There are limits to human abilities to make sense of complex information, and individual scientists may use their own heuristics, and prior experience to collapse and make sense of that information. Those heuristics, and mental shortcuts introduce bias and error into the analysis (Hruschka et al., 2004; MacQueen et al., 1998; Tversky & Kahneman, 1974). One way to reduce this error is to give clear definitions of each procedural unit. 
Clear definitions, however, are not enough. If I gave a careful definition of the procedural unit “core tablet“, there may be many forms that could be accomodated into that definition that I had not taken into account. Likewise, other researchers may have different standards by which they would count a“ core tablet“ as being present in an assemblage. Some may be happy to find flakes that have the features of a core tablet, and count them as core tablets. Others may only count flakes with those features if there are also cores with negative removals consistent with core tablet removals. Others may have very different ideas of what the features of a core tablet are, and slightly different interpretations of those features themselves. The problem with using only definitions, is that it does not give enough guidance on what forms do not meet that definition, and when we should code “core tablet“ as present. Social scientists at the Center for Disease Control encountered similar problems while coding interview transcripts and other texts for qualitative analysis. Their solution was to develop codebooks with explicit inclusion, and exclusion criteria.  

This codebook follows a format developed at the CDC to ensure reliability in coding texts (MacQueen et al., 1998). Each procedural unit has a short, and longer definition (if needed). Additionally, each also has inclusion criteria. These inclusion criteria describe what features need to be present in a report for us to count the procedural unit as present. Each also has exclusion criteria. These describe what features in the text would be grounds to count the procedural unit as absent. For example, many procedural units only make sense to count as present if hierarchical core reduction is present. So, one exclusion criteria for those procedural units would be the absence of any evidence for hierarchical core reduction.  Each entry may also include typical, and atypical examples of the procedural unit. These may be illustrations, or text examples. These two help to further outline when it is appropriate to code a procedural unit as present. Finally, there is a “close, but no“ entry, which outlines what kinds of cases might be confused with the procedural unit. Some of these inclusion criteria are more exhaustive than others, especially in cases where I noticed that I came to different conclusions about how a procedural unit should be coded multiple times while double checking data. 

Whether or not you agree with all the definitions, and coding criteria, you should be able to come closer to the same conclusions about how to code the presence or absence of any of these procedural units if you follow the standards outlined in the codebook than would be the case without reporting the criteria I employed. The codebook should evolve over time, as new logical inconsistencies or points of confusion are discovered. However, whatever version was used to code the data that then ends up in a final analysis, should be archived and made available to readers. This version of the codebook does not have example illustrations for procedural units to avoid copyright issues, but references to the illustrations are retained.  

# Definitions

**Hierarchical cores**
Cores wherin platforms are established to strike flakes that shape the main flaking surface of a core, or the main platform itself in preparation for removals across the main flaking surface. 


**Main flaking axis** 
On hierarchical cores, this is the axis parallel to the face along which most target pieces were produced. On a naviform core, for example, the main flaking axis is that along which bidirectional blades are taken. Radial cores have no main flaking axis. Centripetal levallois cores also have no main falking axis, while preferential levallois cores do. 


**Main flaking surface** 
On hierarchical cores, this is the surface from which targetted blanks were removed.


**Core platform** 
Any of the edges percussed or applied with pressure to remove flakes. 


**Main core platform(s)**
The platforms used to produce target pieces. Blade platforms on blade cores. Facetted platform on a preferential levallois. 


**Core back**
Posterior surface relative to the main flaking surface, roughly parallel to the main flaking surface.


**Core bottom**
Distal surface of a core, opposite the main core platform.  


# Procedural unit codes

## Raw material treatment

**Short Description** Heat treatment of raw material

**Definition** Heating of raw material in order to improve workability. This process alters the fracture mechanics of raw material, and often causes changes in texture, and color. 

**Inclusion criteria** If heat treatment is described as present, code as yes. If heating is described, and reference made to glassy/glossy texture of raw material as a result of heating, code as present.

**Exclusion criteria** Heat treatment not mentioned in text. 

**Typical exemplars** “Flint was/was likely heat treated”

**Atypical exemplars** “flint was glossed/waxy/greasy from heating”

**Close but no** “material was fired”, “material bears signs of thermal alteration”, Images of pieces that appear to have been heat treated, but without accompanying text describing them as heat treated. “Some flints had a glossy/waxy/greasy appearance”.


## Faceting of core platform

**Short Description** Shaping of a core platform by striking flakes into the platform from the face of the core. 

**Definition** Removal of two or more flakes struck into the face of a core across the platform, forming a platform with two or more parallel facets.

**Inclusion criteria** Include in cases only where hierarchical core reduction is described. Code if facetting of platforms is described in the text as a method of preparing platforms. Code if illustrations show evidence of faceting. 

**Exclusion criteria** Determining presence based on illustrations alone requires illustrations of the platforms themselves. Inferring platform faceting from dorsal or ventral views of a flake are inappropriate. Do not code if no mention of faceting of platforms.  Do not code if no other evidence of hierarchical core reduction described. Do not code if the flakes were likely not struck into the face of the core

**Typical exemplars** Figure 1. Pieces a, c, d, e, f, g, h. 

**Atypical exemplars** Figure 1. Piece b. Figure 2. Pieces 1, 3-7, Figure 3. Step D. 

**Close but no**  Phrases like “platforms were carefully prepared” without additional supporting information about the nature of that preparation. Single or very sparse instances of facetted or dihedral platforms, especially where no other evidence of hierarchical reduction** “Nevertheless it is important to mention that within the excluded material of the ‘MSA base-complex’ (see above) one flake… shows characteristics of a Levallois preferential flake with centripetal dorsal scars and a facetted striking platform...” (Schmidt, 2011, p. 92). All examples in figure 4 would be insufficient to code as present. 


## Face shaping through radial removals

**Short Description** Shaping of the face of a core through centripetal removals along the perimeter of a core face. 

**Definition** Flakes taken to modify the distal, lateral, medial convexities of a round core face, to prepare it for preferential removals. The preferential removals could be unidirectional or bidirectional blades, or preferential flakes. 

**Inclusion criteria** Consistent evidence for radial scars on blanks. Cores with evidence of centripetal preparation and hierarchical setup. Core faces should be rounded/oval, not rectilinear. Description of centripetal preparation of cores. 

**Exclusion criteria** radial cores without evidence for preferential removals, presence of only lateral trimming, or distal trimming flakes on a prepared core. 

**Typical exemplars** Figure 5. Pieces 1, 2, 4, 7.

**Atypical exemplars**

**Close but no**  Figure 5. Pieces 3, 5, 6. Discoid cores, non-hierarchical core faces, biface thinning, cores of a rectilinear shape with both lateral trimming and distal trimming.  


## Lateral trimming

**Short Description** Lateral convexity of core face is shaped with flakes initiated from platforms along lateral margins of core.  
Definition** The lateral convexities of the face of a core are trimmed through the removal of flakes from the lateral margin of the face (i.e. from platforms parallel to main flaking axis) 

**Inclusion criteria** Explicit descriptions of lateral trimming on hierarchical cores, and unambiguous illustrations of cores with lateral trimming. 

**Exclusion criteria** Do not count if** lateral flakes were struck only during radial/centripetal preparation of a core face, If lateral convexities are only trimmed through debordante removals, or if part of lateral margin of a core has flake removals on it, but that core is not preferential, or hierarchical. Also do not count if the lateral margin of a core has flake scars perpendicular to the main flaking axis, but those flakes originated from a crest used to establish the core face, or otherwise did not originate from a platform at the lateral margin of the core.

**Typical exemplars** Figure 6, no. 4.

**Atypical exemplars**  Figure 5. Piece 3. 

**Close but no** descriptions of debordant, but without further specifics about the orientation of trimming flakes.
 

## Distal trimming

**Short Description** Face shaped with flakes initiated from platform at distal margin of core
Definition** Distal convexities of the face of a hierarchical core are managed through the removal of flakes from a platform at the distal margin of the core (i.e. the platform is perpendicular to the main flaking axis).

**Inclusion criteria** Core faces show evidence for flake removals from distal margin. Must be on hierarchical cores, and must be in context of managing distal convexities of the core. 

**Exclusion criteria** Do not count if** the core could be reasonably characterized as centripetally prepared, if these are flakes initiated from the distal margin that trim the lateral margins of a core, like what we might find in a Nubian Levallois core. Do not count as present if no hierarchical cores are present. 

**Typical exemplars** Figure 5, Levallois cores numbers 5, 6., Figure 6, Levallois core with lateral trimming, number 4. Figure 7. Step 4 in blademaking sequence.

**Atypical exemplars** Figure 8. Step 3 in Chazan’s description of La Ferrassie bladelets. Notching a bladelet, to supply end point of microblades taken from lateral margin of that bladelet core. Controls length of flaking surface by establishing/shaping a distal convexity.

**Close but no** 


## Back shaping

**Short Description** Back of the core is shaped.

**Definition** Back of core is shaped, as is case among naviform cores, and Asian microblade cores. All examples so far identified are cases where a nodule was bifacially flaked. One of the flaked crests then is used to remove one or two crested blades to establish a platform and face. Then flakes are removed from that platfom until exhausted. When exhausted there remains evidence of original bifacial flaking at the back of the core. 

**Inclusion criteria** Back of hierarchical core is shaped, typically bifacially. 

**Exclusion criteria** Crested blades are present, but no illustrations of cores showing evidence for a modified back/non-flaking surface. 

**Typical exemplars** Figure 9. Piece 2b. 

**Atypical exemplars** NA 

**Close but no**  NA


## Cresting

**Short Description** Cresting to shape core face during initial steps of core preparation. 
Definition** A core is bifacially or unifacially flaked along one axis. That crest establishes an artificial ridge along which an elongated flake, with a crested (entirely or partially) dorsal surface is removed. 

**Inclusion criteria** Cresting of core faces is described as present and/or figures show elements flakes bearing a crest, with at least a partial crest platform present. 

**Exclusion criteria**  Cresting of core faces is described as absent. No mention in the text, and there are no illustration of crested pieces. For example** from Shimelmitz et. Al. 2011** “The reduction takes advantage of the natural shape of the raw material and does not include pre-shaping and decortication”. 

**Typical exemplars** Figure 10. Flakes whose dorsal surface bears a full bifacially flaked crest. Pieces 1, 2, 4, 5, 6, 7.

**Atypical exemplars** From Wilkins and Chazan** “The rarity of crested blades (0.3%, Table 7) and the presence of blades with centripetal flake dorsal scars on one side only (4.2%, Fig. 9e,h) are most consistent with a reduction strategy that generally prepared the blade exploitation surface with centripetal flake removals”(Wilkins & Chazan, 2012** p. 10). This is enough to code cresting as present in the assemblage. “Crested elements (n=26) provide some details of the methods used for initial core preparation. The relatively low number of these pieces compared to the high number of cores used to produce blade/bladelet suggests that, in general, cresting was not necessary, and the natural shape of cobbles/nodules allowed the production of elongated debitage without cresting.”(Smith et al., 2016).

**Close but no** flakes with laterally oriented dorsal scars. Partially crested blades, where no element of the platform of the crest is present. Ski spall flakes removed during reduction of naviform cores which were prepared through cresting. Flakes with unifacial cresting, either complete or partial as in the case of striking platform removals as described in Smith et al. 2016.


## Debordantes

**Short Description** Elongated flakes along lateral margins of core face, knapped along the axis of the main flaking surface.  

**Definition** Elongated flakes removed from lateral margins of core face, knapped along the axis of the main flaking surface. Debordantes only include materials that maintain lateral and distal and sometimes proximal convexities of a core face. These can be removed either from the proximal/main core platform (more common), or distal area of the core (see **Atypical exemplars below). These tend to have a triangular cross section. In near eastern traditions these are sometimes refered to as naturally backed blades or knives (Shimelmitz et al., 2011; Smith et al., 2016) though these sometimes might refer to elements that are not from hierarchical preferential cores, that happen to have a wedge shaped cross section. 

**Inclusion criteria**  Must be in association with description/illustration of hierarchical preferential cores. If that condition is met, and the term debordante is used and likely is not referring to what we would otherwise code as lateral trimming, count as present. If backed flake/knife is used, only count as present if there is further discussion about the role they play in managing convexities of core face, OR if core faces show clear evidence of debordante removals in illustrations. If “debordante” is not used, but there are phrases like “elongated flakes were used to alighn the core face/modify lateral and distal convexities/modify core edges” count as present. 

**Exclusion criteria** Naturally backed flakes described as present, but there is no mention of their function in maintaining the face geometry of the core, or convexities of the core. Flakes are described as debordante, but there is otherwise no mention illustration or description of the nature of cores. Or, there is description of cores and they are non-hierarchical/amorphous. 

**Typical exemplars** From Wilkins and Chazan** “Débordant flakes/blades with unidirectional or bidirectional dorsal scars and a preserved lateral platform surface are also present in the assemblage (n ¼ 13, 1.3%, Fig. 9a) and may have sometimes been used to rejuvenate core edges”(Wilkins & Chazan, 2012). Figure 5. Pieces 5, 6, and 9 in figure below.

**Atypical exemplars** Naturally backed blades described in Shimelmitz et al. 2011. Pieces that are described as backed but in context of discussion about managing lateral convexities. For example** “Only one elongated flake could be attributed to the Taramsa reduction method. The presence of many backed pieces (N=10) confirms this trend, since these flakes are removed to maintain the strong lateral convexity of the core”(Spinapolice & Garcea, 2013). 
Figure 5. Piece 8. 

**Close but no** Discussion of face shaping with flake removals or discussion of naturally backed blades/knives/flakes without mention of the role they served in maintaining face shape/convexity. 


## Overshot flakes

**Short Description** Elongated flake removals that clip or remove the distal margin of the core. 

**Definition** Medial and distal convexities of a core face on a preferential hierarchical core are modified with an invasive flake that removes the distal end of the core, which may bear a platform (often opposed to its own) on the distal margin. 

**Inclusion criteria** Flakes or blades that extend across the entire face of the core must be in association with hierarchical blade/bladelet/microblade cores, discussion must include discussion of overshot blades/flakes, and/or include illustrations of overshot flakes/blades. 

**Exclusion criteria**  No argument for presence. No Hierarchical preferential cores. No overshot blades/flakes in illustrations. No discussion of overshot blade/flake role in modifying distal convexity/rejuvenating distal platform. 

**Typical exemplars**  Shimelmitz et al. 2011** “The frequent removal of laminar items with an overpassing end termination along the reduction in order to control core convexities.”

**Atypical exemplars** From Wilkins and Chazan. “Blades preserving a distal striking platform (Fig. 7d) further attest to the bidirectional production of blades.”. This phrase has enough information for us to code this as an overshot blade, but there is also supporting information in the figure showing overshot blades. 

**Close but no**  Blades that are thick at the distal end, but do not bear the distal end of the core. Flakes that have a distal end of core, but the distal end is cortex. Overshot flakes are present in an assemblage, but there is no evidence for hierarchical core reduction. 

## Kombewa

**Short Description** Removal of flake from ventral surface of a flake	

**Definition** Ventral surface of a flake is treated as a core face

**Inclusion criteria** Kombewa technique described as present, “janus flakes” and the processes to make them are described

**Exclusion criteria** Discussion of removal of flakes from flakes is ambiguous about whether or not kombewa technique is present. No mention of janus flakes. 

**Typical exemplars** “cleavers were made on sometimes completely unmodified Kombewa flakes”

**Atypical exemplars** 

**Close but no**  Core on flake described as present without discussion of where the flaking surface is on that flake. Flakes from retouching large flakes, which be initiated at dorsal margin, and capture some of the ventral surface. Scars not propagated across ventral surface. Burin spalls, or tranchet spalls taken from flakes.


## Core tablet removals

**Short Description** Flake removals that rejuvinate or prepare a core platform, by removing some or all of the core platform. 

**Definition** Striking a flake into the face of a core, where the dorsal surface of that flake is the main platform of a core. These are intended to rejuvinate the core platform by establishing a fresh flaking surface. 

**Inclusion criteria** Descriptions of core tablets, or illustrations of core tablets themselves and the function they served in rejuvenating core platforms, or illustrations of cores with strong evidence for core tablet removals. 

**Exclusion criteria** Coding as present should not be based only on illustrations of pieces that look like tablets. It should also not normally be based on illustrations of cores that have a single scar as the platform without supporting evidence that that scar was made to rejuvenate the platfrom (see Atypical exemplars below). Do not code as present if only evidences are flakes that happen to remove platforms, but no other evidence for hierarchical cores, Flakes with large facetted proximal margins without any other information in the text about rejuvenation of platforms. 

**Typical exemplars** Figure 11. Cores a and d. Phrases like, ‘the assemblage has several core tablets’.

**Atypical exemplars** Figure 11. Cores b and e. In Figure 12 we can see that a flake was taken across the top of the core (the negative bulb of the flake is present) and this flake erased the negative bulbs of several of the blade scars (the flake did not just establish a platform, but was taken after several blades had been struck. These lines of evidence on a single platform hierarchical core are sufficient to infer that these cores were rejuvenated with tablets. 

**Close but no** Figure 12. Pieces 1-4. Without additional information about the geometry of the core from which these were taken, we should not call these core tablets (though the paper from which the figure is borrowed provides enough context to call these tablets). 


## Abrasion/grinding 

**Short Description** Abrasion or grinding performed at any point in reduction sequence.

**Definition** Core was abraded/ground to strengthen platform, or tools were abraded or ground as part of production sequence. 

**Inclusion criteria** any discussion of the abrasion, rubbing, of the core platform in relation to platform preparation in the text, or the presence of grinding/abrasion to finish a tool.

**Exclusion criteria** No explicit description of abrasion as the strategy used to prepare the platform. Platform preparation described, but the kind of preparation not explicitly described. Tools appear ground/abraded but no discussion of the technique in text. 

**Typical exemplars**  ‘platforms were prepared by abrasion.’

**Atypical exemplars** “These bifaces have several common attributes. All were primarily shaped by abrasion by rubbing with a coarse material, leaving parallel striae on their surface” (Rosenthal, 1996). 

**Close but no**  NA


## Overhang removal/microchipping of area below platform

**Short Description** Removal of chips to modify area below platform. 

**Definition** Removal of chips initiated from platform to modify proximal margin of core face/proximal convexities, and modify the platform angle. 

**Inclusion criteria** Illustrations show small flakes removed at proximal margin of flakes, or on areas below core platforms. Descriptions of overhang removal, or microchipping of platform in text. 

**Exclusion criteria**  No evidence of microchipping, overhang removal described in text, or in illustrations. Microchipping described in text, but this refers to what we would otherwise code as faceting. 

**Typical exemplars** NA

**Atypical exemplars** NA 

**Close but no**  NA


## Percussion by striking with hard hammer

**Short Description** Use of a hard hammer

**Definition**  Use of a hard hammer strike onto some substrate, whether it is a core, held in the had, mounted on an anvil, or whether the hammer itself was struck on an anvil (as in case of passive hammer technique)

**Inclusion criteria**  If flakes are produced in assemblage, count as present unless explicitly stated as absent

**Exclusion criteria** Explicit statement in text saying hard hammer use was absent

**Typical exemplars** NA

**Atypical exemplars** NA

**Close but no**  NA


## Core supported by hand

**Short Description** Any stage of tool manufacture includes holding the core in hand while striking it (i.e. use of anvil is absent). 

**Definition**  NA

**Inclusion criteria**  Count as present if bipolar percussion is not described as present, if description of freehand percussion in text. 

**Exclusion criteria** Explicit statement in text saying only bipolar, or passive anvil technique was employed, or description only of flake removals while core was mechanically mounted, or otherwise not supported by hand.

**Typical exemplars** NA

**Atypical exemplars** NA

**Close but no**  NA


## Use of anvil to support core

**Short Description** Any incorporation of an anvil in the reduction process. 

**Definition** Use of an anvil at any point in the tool reduction sequence. 

**Inclusion criteria** Any discussion of anvil use, all cases of bipolar percussion, except where bipolar refers to flake removals from two opposing platforms where an anvil is not used (i.e. a bidirectional core without bipolar percussion). Cases where bipolar is mentioned must also have some visual evidence for bipolar percussion with an anvil (in the form of scaled pieces, for example).

**Exclusion criteria** Bipolar percussion not mentioned in text. No mention of use of anvil to support core in any way.

**Typical exemplars** “…many of our early replications were performed with the aid of an anvil (Figure 5), and this was found to be a successful technique for creating the initial steep sides on large flakes and cobbles. Anvil resting was more successful than true bipolar flaking in generating steep-angled edges and moving flaking on to new edges” (Clarkson et al., 2015** p.74)

**Atypical exemplars** NA


## Core rotation

**Definition** Core rotated at any point in reduction sequence

**Inclusion criteria** Any evidence of flakes removed across two or more distinct axes. 

**Exclusion criteria** Single platform cores are present without evidence of removals across the top of the core. 

**Typical exemplars** NA

**Atypical exemplars** NA

**Close but no** NA


## Soft Hammer

**Short Description** Use of a soft hammer 

**Definition** use of a soft hammer (whether the material be wood, bone, soft stone, etc). 

**Inclusion criteria** Authors explicitly state that soft hammer percussion was likely used OR soft hammers present in the archaeological assemblage, and forms of pieces appear consistent with use of soft hammers. For example, if delicate, thin, wide flakes present and soft-hammers were recovered from related contexts. 

**Exclusion criteria** No mention of soft hammer use, no soft-hammers present in the assemblage.  Evidence for soft hammer use described as unclear. 

**Typical exemplars** ‘Soft hammer use likely occurred’

**Atypical exemplars** ‘Soft hammers are necessary for producing these forms, also see Figure x. for soft hammers recovered from the archaeological record’

**Close but no**  ‘Soft hammer use could have been employed/other assemblages with similar forms have soft hammers’


## Indirect percussion

**Short Description** Use of a punch to remove flakes

**Definition** Use of a punch of any given material placed on a platform, and struck with a hammer to punch flakes from the core. 

**Inclusion criteria** Authors explicitly state that indirect percussion was likely used.

**Exclusion criteria** No mention of indirect percussion, statements like “indirect percussion is one method that could produce the forms here”.

**Typical exemplars** ‘The blades in this assemblages would have likely required indirect percussion’, ‘experimental reconstructions of the stitching pattern on these Danish daggers indicate that indirect percussion/use of a punch would have been required’

**Atypical exemplars** NA

**Close but no**  ‘Indirect percussion could have been employed’


## Flaking with application of pressure

**Short Description** Removal of flakes through application of pressure on core platform

**Definition** Use of typically soft indentor, bone, metal, or hard wood, to press flakes off cores.

**Inclusion criteria** Discussion of pressure flaking as means of producing flakes.

**Exclusion criteria** no evidence of scars consistent in pressure flaking AND no discussion of pressure flaking in the text. 

**Typical exemplars** ‘Microblades were struck through application of pressure’

**Atypical exemplars** 

**Close but no**  ‘Other assemblages with similar forms have pressure flakers’


## Pecking/hammer dressing

**Short Description** Modification of core or tool through pecking

**Definition** NA

**Inclusion criteria** Hammer dressing, or pecking described in text as method employed at any point in tool manufacture. Hammer dressing is unambiguously present in illustrations. 

**Exclusion criteria** No hammer dressing or pecking described in text. Illustrations show no unambiguous presence of hammer dressing. 

**Typical exemplars** Figure 13. 

**Atypical exemplars** NA

**Close but no**  NA


## Invasive flaking

**Short Description** Removal of non-cortical flakes that extend beyond the midpoint of the core face

**Definition** NA

**Inclusion criteria** Examples of invasive negative flake scars present in core illustrations. 

**Exclusion criteria** No illustrations of invasive negative flake scars on cores, no description of flakes invasive to the degree that 
they extend beyond midline. Invasive flakes are cortical.

**Typical exemplars** 

**Atypical exemplars** 

**Close but no**  Invasive flaking described only in text, but without further information about how invasive the flake are. 

## Ochre use

**Short Description** Use of ochre in any stage of tool making

**Definition**Use of ochre as a pigment or as a binding agent.

**Inclusion criteria** Ochre applied to tool.

**Exclusion criteria** No mention of ochre in text, and 

**Typical exemplars**  ‘Ochre was applied to points’, ‘the adhesive residues include ochre’

**Atypical exemplars** NA

**Close but no**  NA


## Asphalt use

**Short Description** Use of aslphalt at any stage of the tool making process.

**Definition** Use of asphalt as a binding agent

**Inclusion criteria** Asphalt adhered to tool, typically at its base/tang.

**Exclusion criteria** No mention of asphalt residue on tools in the text. 

**Typical exemplars**  ‘asphalt was applied to points’, ‘the adhesive residues include asphalt’

**Atypical exemplars** NA

**Close but no**  NA


## Tanging

**Short Description** Retouching base of piece to form a tang

**Definition** Retouching a piece, typically through backing and notching at the base of a piece in order to facilitate hafting the piece. 

**Inclusion criteria** Description of retouch as forming a tang for the purpose of hafting. 

**Exclusion criteria** No mention of tanging, or hafting. Pieces bearing notches are not described as hafted. 

**Typical exemplars** NA

**Atypical exemplars** No mention of tanging, but pieces with basal modifications consistent with a tang are described as frequent or have many illustrations. 

**Close but no**  NA


## Invasive retouch

**Short Description** retouch that extends to the midline of the artifact. 

**Definition** NA

**Inclusion criteria** Illustrations of retouched pieces showing retouch extending to midline of tool.  

**Exclusion criteria** No illustrations of retouched pieces. 

**Typical exemplars**  

**Atypical exemplars** Burin spalls that extend to the midline of the artifact. Tranchet spalls. 

**Close but no**  Retouch described as invasive, but retouch scars in illustrations extend short of the midline of the artifact. 


## Retouch (unifacial)

**Short Description** Retouch of flake (unifacial only)

**Definition** 

**Inclusion criteria** Retouch described as present. Illustrations of pieces show unambiguous evidence of retouch. 

**Exclusion criteria** Bifacial retouch is present. No unambiguous illustrations of unifacially retouched pieces, and no mention of 
retouch. 

**Typical exemplars** NA

**Atypical exemplars** NA

**Close but no**  NA


## Backing

**Short Description** Retouch forms an abrupt, scraper-like margin

**Definition** Retouch that increases the angle of the margin to ~70-90 degrees.

**Inclusion criteria** Presence of artifacts with retouch that forms an angle greater than 70 degrees. Description of scrapers in 
assemblage.

**Exclusion criteria** Lack of evidence for artifacts with retouch forming greater than 70 degree angle. Or evidence is ambiguous. No 
mention of scrapers, or illustrations of retouched tools. 

**Typical exemplars** 

**Atypical exemplars** 

**Close but no**  Abrupt retouch on flat thin flakes. 


## Notching

**Short Description** Retouch forms round concavity. 

**Definition** Retouch, either unifacial or bifacial, forms a round concavity, or series of concavities.

**Inclusion criteria** Description of notches, or denticulates. Illustration of pieces with notches formed. 

**Exclusion criteria** Lack of evidence for artifacts with retouch forming greater than 70 degree angle. Or evidence is ambiguous. No 
mention of scrapers, or illustrations of retouched tools. 

**Typical exemplars** Figure 14. Piece 1. Figure 15. Pieces E, F.

**Atypical exemplars** 

**Close but no**  Abrupt retouch on flat thin flakes. 


## Burination

**Short Description** Removal of spalls along the margins of other flakes. 

**Definition** Removal of flakes where the core face is the sharp margin of the flake. Flakes from this process have two ventral surfaces, the parent flake’s, and its own. 

**Inclusion criteria** Illustrations of flakes with evidence of spalls taking across their lateral, proximal, or distal margins. Burins, or burin spalls described as present. 

**Exclusion criteria** No description of burins, burin spalls, or microburin technique, and no illustrations showing burins as described 
above. Do not count if the burination could be coded as tranchet resharpening

**Typical exemplars** Figure 16. from Smith et al. 2016. all examples.

**Atypical exemplars** NA

**Close but no**  Mention of ‘impact burnation’, ‘spalling’, ‘core-on-flake’.


## Tranchet removal.

**Short Description** Retouch of a core-tool by removing a flake across the face at the distal margin or bit.

**Definition** Retouch of a core-tool by removing a flake across the face at the distal margin or bit. The spall removed in this process is typically curved, and may be trihedral in a way similar to a burin spall. One lateral margin of the spall will be the distal margin/bit of the core-tool. The opposite face of the core tool will have a remnant on one face of the spall, adjacent to the ventral surface of the spall. 

**Inclusion criteria** Description of tranchet resharpening in text.  illustration of core-tools with negative from tranchet spall visible. Illustration of tranchet spalls themselves with further description that these represent tranchet spalls. 

**Exclusion criteria** Do not count if there is no evidence in illustrations for tranchet resharpened core tools, and there is no mention of tranchet spall in text, or otherwise no mention of resharpening of bit through removal of transverse flake. 

**Typical exemplars** Figure 17. Pieces 1-3.

**Atypical exemplars** Illustrations of core-tools with unambiguous examples of tranchet resharpening, but no description of this method of resharpening in text. 

**Close but no**  Illustrations of what appear to be tranchet spalls, in context of a site with core-tools that could reasonably have been resharpened with such spalls, but no description in the text that these spalls served that purpose. 


## Pressure retouch

**Short Description** Pressure flaking retouch

**Definition** Retouching a piece (core-tool or flake) with application of pressure. 

**Inclusion criteria** Invasive flake scars on retouched pieces that are extremely narrow (~.5mm), thin (<.1mm, and relatively invasive (~5mm) OR description of pieces as pressure flaked. 

**Exclusion criteria** no evidence of scars consistent in pressure flaking AND no discussion of pressure flaking in the text. 

**Typical exemplars** ‘There were x pressure flaked bifaces in the assemblage’

**Atypical exemplars** ‘Burin spalls removed through pressure flaking’, or ‘microblade manufacture on end nosed scraper were removed through pressure flaking.’

**Close but no**  ‘biface was thinned through invasive and delicate removals’, ‘delicate burin spalls removed’. 


## Bifacial retouch

**Short Description** Retouch on both faces of a flake or core-tool. 

Definition** Retouch on both faces of a flake or core-tool, struck from the same platform. 

**Inclusion criteria** Descriptions of bifacial retouch, illustrations of pieces with retouch on both faces.

**Exclusion criteria** No mention of bifacial retouch, no illustrations of retouched pieces showing bifacial retouch.

**Typical exemplars** NA

**Atypical exemplars** NA

**Close but no**  NA

**Other notes** Never count both bifacial and unifacial retouch on same piece. 

# Figures

 
Figure 1. Figure 23 illustrating levallois point variation from Kibish formation (Shea, 2008). 
 
Figure 2. From figure 5 illustrating levallois flake and centripetal flake diversity (Picin & Vaquero, 2016). 
 
Figure 3. Figure 2 illustrating schematic drawings of blade manufacture methods in Queensland (Moore, 2003).

Figure 4. Figure 4 on technological blade classifications at Rose Cottage Cave (Soriano et al., 2007).
 
Figure 5. Figure 1 in in description of Levallois technology (Bordes, 1980).

Figure 6. Figure 2 in description of Levallois technology (Bordes, 1980). 

Figure 7. Figure 5 description of bladelet core preparation at ‘Ein Qashish with distal preparation at step 4.(Malinsky-Buller et al., 2014) 

Figure 8. Figure 2. Schematic illustrating busqued burin production methods at La Ferrassie (Chazan, 2001).

Figure 9. Figure 2. Microblade core variability at Amakomanak.(Coutouly, 2017)

Figure 10. Initial blade subtypes from Kfar HaHoresh (Barzilai & Goring-Morris, 2010). 

Figure 11. Figure 4 in description of blade cores from Fumane cave (Falcucci & Peresani, 2018).
 
Figure 12. Figure 14 illustrating platform spalls from bidirectional blade cores recovered from Kfar HaHoresh (Barzilai & Goring-Morris, 2010)
 
Figure 13. Figure 2 illustrating hammer dressing on stemmed obsidian tool from Biak Island, West Papua (Robin Torrence et al., 2009).

Figure 14. Figure 13 illustrating projectile points recovered from Motza (Khalaily et al., 2007). 
 
Figure 15. Figure 8 illustrating some retouched tool tyles from Ayn Abu Nukhayla (Henry & Mraz, 2020).

Figure 16. Figure 5 illustrating burin variation at the PPNA site El Hemmeh (Smith et al., 2016).  
 
Figure 17. Figure 14 illustrating tranchet axe variability at Motza (Khalaily et al., 2007).

# Bibliography

x


Barzilai, O., & Goring-Morris, A. N. (2010). Bidirectional Blade Production at the PPNB Site of Kfar HaHoresh: The Techno-Typological Analysis of a Workshop Dump. Paléorient, 36(2), 5–34. https://doi.org/10.3406/paleo.2010.5386

Bordes, F. (1980). Le débitage Levallois et ses variantes. Bulletin de la Société préhistorique française, 77(2), 45–49. https://doi.org/10.3406/bspf.1980.5242

Chazan, M. (2001). Bladelet Production in the Aurignacian of La Ferrassie (Dordogne, France). Lithic Technology, 26(1), 16–28. https://doi.org/10.1080/01977261.2001.11720973

Clarkson, C., Shipton, C., & Weisler, M. (2015). Front, back and sides: Experimental replication and archaeological analysis of Hawaiian adzes and associated debitage: A study of Hawaiian adze manufacture. Archaeology in Oceania, 50(2), 71–84. https://doi.org/10.1002/arco.5056

Coutouly, Y. A. G. (2017). Amakomanak: An Early Holocene Microblade Site in Northwestern Alaska. Arctic Anthropology, 54(2), 111–135. https://doi.org/10.3368/aa.54.2.111

Falcucci, A., & Peresani, M. (2018). Protoaurignacian Core Reduction Procedures: Blade and Bladelet Technologies at Fumane Cave. Lithic Technology, 43(2), 125–140. https://doi.org/10.1080/01977261.2018.1439681

Henry, D. O., & Mraz, V. (2020). Lithic economy and prehistoric human behavioral ecology viewed from southern Jordan. Journal of Archaeological Science: Reports, 29, 102089. https://doi.org/10.1016/j.jasrep.2019.102089

Hruschka, D. J., Schwartz, D., St.John, D. C., Picone-Decaro, E., Jenkins, R. A., & Carey, J. W. (2004). Reliability in Coding Open-Ended Data: Lessons Learned from HIV Behavioral Research. Field Methods, 16(3), 307–331. https://doi.org/10.1177/1525822X04266540

Khalaily, H., Bar-Yosef, O., Boaretto, E., Bocquentin, F., Le Dosseur, G., Erikh-Rose, A., Goring-Morris, A. N., Greenhut, Z., Marder, O., Sapir-Hen, L., & Yizhaq, M. (2007). Excavations at Motza in the Judean Hills and the Early Pre-Pottery Neolithic B in the Southern Levant.pdf. Paleorient, 33(2), 5–37.

MacQueen, K. M., McLellan, E., Kay, K., & Milstein, B. (1998). Codebook Development for Team-Based Qualitative Analysis. CAM Journal, 10(2), 31–36. https://doi.org/10.1177/1525822X980100020301

Malinsky-Buller, A., Ekshtain, R., & Hovers, E. (2014). Organization of lithic technology at ‘Ein Qashish, a late Middle Paleolithic open-air site in Israel. Quaternary International, 331, 234–247. https://doi.org/10.1016/j.quaint.2013.05.004

Moore, M. W. (2003). Australian Aboriginal Blade Production Methods on the Georgina River, Camooweal, Queensland. Lithic Technology, 28(1), 35–63. https://doi.org/10.1080/01977261.2003.11721001

Perreault, C., Brantingham, J., Kuhn, S. L., Wurz, S., & Gao, X. (2013). Measuring the Complexity of Lithic Technology. Current Anthropology, 54(8).

Picin, A., & Vaquero, M. (2016). Flake productivity in the Levallois recurrent centripetal and discoid technologies: New insights from experimental and archaeological lithic series. Journal of Archaeological Science: Reports, 8, 70–81. https://doi.org/10.1016/j.jasrep.2016.05.062


Rosenthal, E. J. (1996). San Nicolas Island Bifaces: A Distinctive Stone Tool Manufacturing Technique. Journal of California and Great Basin Archaeology, 18(2).


Schmidt, I. (2011). A Middle Stone Age Assemblage with Discoid Lithic Technology from Etemba 14, Erongo Mountains, Northern Namibia. Journal of African Archaeology, 9(1), 85–100.

Shea, J. J. (2008). The Middle Stone Age archaeology of the Lower Omo Valley Kibish Formation: Excavations, lithic assemblages, and inferred patterns of early Homo sapiens behavior. Journal of Human Evolution, 55(3), 448–485. https://doi.org/10.1016/j.jhevol.2008.05.014

Shimelmitz, R., Barkai, R., & Gopher, A. (2011). Systematic blade production at late Lower Paleolithic (400–200 kyr) Qesem Cave, Israel. Journal of Human Evolution, 61(4), 458–479. https://doi.org/10.1016/j.jhevol.2011.06.003

Smith, S., Paige, J., & Makarewicz, C. A. (2016). Further diversity in the Early Neolithic of the Southern Levant: A first look at the PPNA chipped stone tool assemblage from el-Hemmeh, Southern Jordan. Paléorient, 42(1), 7–25. https://doi.org/10.3406/paleo.2016.5691

Soriano, S., Villa, P., & Wadley, L. (2007). Blade technology and tool forms in the Middle Stone Age of South Africa: The Howiesons Poort and post-Howiesons Poort at Rose Cottage Cave. Journal of Archaeological Science, 34(5), 681–703. https://doi.org/10.1016/j.jas.2006.06.017

Spinapolice, E. E., & Garcea, E. A. A. (2013). The Aterian from the Jebel Gharbi (Libya): New Technological Perspectives from North Africa. African Archaeological Review, 30(2), 169–194. https://doi.org/10.1007/s10437-013-9135-2

Tversky, A., & Kahneman, D. (1974). Judgment under Uncertainty: Heuristics and Biases. Science, 185(4157), 1124–1131. https://doi.org/10.1126/science.185.4157.1124

Wilkins, J., & Chazan, M. (2012). Blade production ~500 thousand years ago at Kathu Pan 1, South Africa support for a multiple origins hypothesis. Journal of Archaeological Science, 1–18.


